Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-14 paper.tex

%\comment{This is the old equation; I think the $e^{-rs}$ should not be there,
%    since it's already taken into account by \eqref{eq:span}.
%    Plus, the $\lambda$ and $r$ factors are now elsewhere.}
%\begin{equation}\label{eq:depth_old}
%B_{(c, p)}(\theta) = \begin{cases}
%e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
%\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c_s, s, t_{p})} ds \lambda^{I_{c_{s}d}(t_p)} & x=x_{c_{s}0}>x_{c0} \\
%\end{cases}
%\end{equation}
%\comment{Do we still need this bit?}
%Here, the (second) exponential term gives the probability of not observing a
%coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further
%gives the probability of observing a recombination
%at time $s$ given a Poisson process along position $x$ with rate $r$.
%The time of the recombination is integrated out.
%The last term gives the point probability density of the coalescence event
%that terminates the edge. The indicator function is used to resolve
%the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.
The overall likelihood of an ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ is then given by
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|r, \lambda)
    =
    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(p, c) \in E} A_e(r) B_e(r, \lambda) .
\end{equation}
% One significant advantage of this formulation is that the computation time
% is linear in the number of edges, and an algorithm can be devised to 
% to efficiently keep track of $I_e(t)$ (see Appendix).
% Although the first two factors are easily expressed in terms of numbers of edges and nodes,
% we will see below
% that in practice it is convenient to absorb these terms into per-edge contributions.
% We can also accommodate stepwise changes in the 
% coalescence rate (or recombination rate)
% through time by intersecting the internode intervals with the rate change
% intervals; rate changes along the genome can be incorporated similarly.
Note that we have only considered the topology and branch length information here
an not considered mutational information.
The likelihood of a set of mutations given an ARG is
\begin{equation}\label{eq:lik-mut}
    \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
    =
    \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
\end{equation}
where $\mathcal{M}$ the total number of mutations and $m_e$
the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.

% % % % % % % % % % % % % % % % % %
\subsection{Flexibility}
% likelihood surface
The last paragraph aims to highlight the flexibility of this approach 
and the impact of the various
degrees of completeness of the ARG on the likelihood.
Note that the figures in this paragraph only serve this exact purpose,
we are not claiming to have designed a new parameter estimation approach.

In line with the observation that the presence of locally unary nodes 
(see section \ref{par:recording}) is crucial for the equivalence between
the graph or local trees view of an ARG, one can gradually simplify down
a simulated ARG by gradually removing the various types of locally unary nodes.
Figure \ref{fig:lik-surface} shows the impact of these simplification steps
on the likelihood for a single simulated ARG.
The log likelihood ratio is shown to improve comparability.

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{figures/likratio.png}
    \caption{Log likelihood-ratio curves for ARGs simulated under the SMC with
    various degrees of completeness. The likelihood ratio is computed with
    respect to the true recombination rate. In case of the full ARG the likelihood
    under the SMC has been computed using the exact time to the recombination event.
    Simulation performed with \texttt{msprime}, parameters: $L=1$Mb, $20$ diploid samples,
    $r=2.5e-8$, $N_e=1e4$.}
    \label{fig:lik-surface}
\end{figure}

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
    \caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
    for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
    (5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
    $r=2.5e-8$, $148$ edges in ARG. A single $N_e$ value was estimated for each
    slice of the ARG as defined by the true population size change time points.
    The inferred posterior distribution (right) using the SMC-likelihood was
    obtained using a uniform prior. Number of edges per time slice
    (from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
    omitted for clarity.}
    \label{fig:3-arg-slices}
\end{figure}

\section{Discussion}

Here, we have introduced a scalable algorithm to compute likelihoods
under the SMC for a general class of ARGs. By formulating the SMC backwards in time
we were able to associate a single likelihood value with each edge in the ARG.
The compatibility of the algorithm with the \texttt{tskit} library allows us to
exploit its associated efficiencies both for the current work and its future
applications. This work will contribute significantly
to ARG-based analysis becoming a standard part of the population
and statistical genetics toolkit.
In particular, we see two main applications for the work presented here.
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic
model and an associated set of parameters, will help bridge the gap between the tremendous recent
advances in ARG inference and the current state of the art statistical inference of
past demography.
The second and probably the most important application of the ability to associate a more
general class of ARGs to a principled model is the ability to boost the current efforts
in ARG inference. 
% great starting point for MCMC
Any inferred ARG can be used a good starting point
for a future MCMC-sampler. This potentially implies a massive runtime reduction by
reducing the necessary burn-in period. Furthermore, by integrating out the exact time
to the recombination event we further restrict the size of the ARG space that needs to
be explored. This MCMC-sampler would again rely on the tree sequence data structure to
efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
% hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
To further improve scalability, hybrid ARG-inference methods could be designed which
rely on heuristic approaches to bring the ARG down from the initial pedigree-dominated phase,
and subsequently use a principled approach to perform ARG inference for the second phase.
This idea resembles the tried and tested approach of hybrid backwards-in-time simulations 
that combine Wright-Fisher dynamics in the very recent past and coalescent simulations 
for the remaining time \citep{bhaskar_distortion_2014, nelson_accounting_2020}. This hybrid
approach has enabled large scale simulations while avoiding the documented biases of the
coalescent relative to the Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. 
The same approach could be thus used for ARG-inference where one can rely on a heuristic 
inference method for the recent past, while using a model-based approach beyond that cutoff point.

@@

[TODO another paragraph summarising what you actually do]

>>>>

%\comment{This is the old equation; I think the $e^{-rs}$ should not be there,
%    since it's already taken into account by \eqref{eq:span}.
%    Plus, the $\lambda$ and $r$ factors are now elsewhere.}
%\begin{equation}\label{eq:depth_old}
%B_{(c, p)}(\theta) = \begin{cases}
%e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
%\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c_s, s, t_{p})} ds \lambda^{I_{c_{s}d}(t_p)} & x=x_{c_{s}0}>x_{c0} \\
%\end{cases}
%\end{equation}
%\comment{Do we still need this bit?}
%Here, the (second) exponential term gives the probability of not observing a
%coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further
%gives the probability of observing a recombination
%at time $s$ given a Poisson process along position $x$ with rate $r$.
%The time of the recombination is integrated out.
%The last term gives the point probability density of the coalescence event
%that terminates the edge. The indicator function is used to resolve
%the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.
The overall likelihood of an ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ is then given by
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|r, \lambda)
    =
    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(p, c) \in E} A_e(r) B_e(r, \lambda) .
\end{equation}
One significant advantage of this formulation is that the computation time
is linear in the number of edges, and an efficient algorithm exists
to keep track of $I_e(t)$ (see Appendix).
% Although the first two factors are easily expressed in terms of numbers of edges and nodes,
% we will see below
% that in practice it is convenient to absorb these terms into per-edge contributions.
% We can also accommodate stepwise changes in the 
% coalescence rate (or recombination rate)
% through time by intersecting the internode intervals with the rate change
% intervals; rate changes along the genome can be incorporated similarly.
Note that we have only considered the topology and branch length information here
an not considered mutational information.
The likelihood of a set of mutations given an ARG is
\begin{equation}\label{eq:lik-mut}
    \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
    =
    \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
\end{equation}
where $\mathcal{M}$ the total number of mutations and $m_e$
the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.

% % % % % % % % % % % % % % % % % %
\subsection*{Flexibility}
% likelihood surface
% The last paragraph aims to highlight the flexibility of this approach 
% and the impact of the various
% degrees of completeness of the ARG on the likelihood.
% Note that the figures in this paragraph only serve this exact purpose,
% we are not claiming to have designed a new parameter estimation approach.
In line with the observation that the presence of locally unary nodes 
(see section \ref{par:recording}) is crucial for the equivalence between
the graph or local trees view of an ARG, one can gradually simplify down
a simulated ARG by gradually removing the various types of locally unary nodes.
Figure \ref{fig:lik-surface} shows the impact of these simplification steps
on the likelihood for a single simulated ARG.
The log likelihood ratio is shown to improve comparability.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/likratio.png}
    \caption{Log likelihood-ratio curves for ARGs simulated under the SMC with
    various degrees of completeness. The likelihood ratio is computed with
    respect to the true recombination rate. In case of the full ARG the likelihood
    under the SMC has been computed using the exact time to the recombination event.
    Simulation performed with \texttt{msprime}, parameters: $L=1$Mb, $20$ diploid samples,
    $r=2.5e-8$, $N_e=1e4$.}
    \label{fig:lik-surface}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
    \caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
    for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
    (5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
    $r=2.5e-8$, $148$ edges in ARG. A single $N_e$ value was estimated for each
    slice of the ARG as defined by the true population size change time points.
    The inferred posterior distribution (right) using the SMC-likelihood was
    obtained using a uniform prior. Number of edges per time slice
    (from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
    omitted for clarity.}
    \label{fig:3-arg-slices}
\end{figure}

\section*{Discussion}
[JK: this needs significant work]
Here, we have introduced a scalable algorithm to compute likelihoods
under the SMC for a general class of ARGs. By formulating the SMC backwards in time
we were able to associate a single likelihood value with each edge in the ARG.
The compatibility of the algorithm with the \texttt{tskit} library allows us to
exploit its associated efficiencies both for the current work and its future
applications. This work will contribute significantly
to ARG-based analysis becoming a standard part of the population
and statistical genetics toolkit.
In particular, we see two main applications for the work presented here.
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic
model and an associated set of parameters, will help bridge the gap between the tremendous recent
advances in ARG inference and the current state of the art statistical inference of
past demography.
The second and probably the most important application of the ability to associate a more
general class of ARGs to a principled model is the ability to boost the current efforts
in ARG inference. 
% great starting point for MCMC
Any inferred ARG can be used a good starting point
for a future MCMC-sampler. This potentially implies a massive runtime reduction by
reducing the necessary burn-in period. Furthermore, by integrating out the exact time
to the recombination event we further restrict the size of the ARG space that needs to
be explored. This MCMC-sampler would again rely on the tree sequence data structure to
efficiently generate and evaluate any new proposal (see \citep{mahmoudi_bayesian_2022}).
% hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
To further improve scalability, hybrid ARG-inference methods could be designed which
rely on heuristic approaches to bring the ARG down from the initial pedigree-dominated phase,
and subsequently use a principled approach to perform ARG inference for the second phase.
This idea resembles the tried and tested approach of hybrid backwards-in-time simulations 
that combine Wright-Fisher dynamics in the very recent past and coalescent simulations 
for the remaining time \citep{bhaskar_distortion_2014, nelson_accounting_2020}. This hybrid
approach has enabled large scale simulations while avoiding the documented biases of the
coalescent relative to the Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. 
The same approach could be thus used for ARG-inference where one can rely on a heuristic 
inference method for the recent past, while using a model-based approach beyond that cutoff point.

@@

% [TODO another paragraph summarising what you actually do]
